---
title: "What Riri got Right: Gender & Pay Inequity"
author: "Jaih Hunter-Hill & Katie Harris"
date: 'December 14, 2020'
output: html_document
---

```{r, echo=FALSE}
library(tidyverse)
library(knitr)
library(GGally)
```

```{r, echo=FALSE}
# Import starting data
nlsy <- read_csv("http://www.andrew.cmu.edu/user/achoulde/94842/final_project/nlsy97/nlsy97_Nov2020.csv")

encoded.names <- names(nlsy)

# Change column names to question name abbreviations (you will want to change these further)
colnames(nlsy) <- c("PSTRAN_GPA.01_PSTR",
    "INCARC_TOTNUM_XRND",
    "INCARC_AGE_FIRST_XRND",
    "INCARC_LENGTH_LONGEST_XRND",
    "PUBID_1997",
    "YSCH-36400_1997",
    "YSCH-37000_1997",
    "YSAQ-010_1997",
    "YSAQ-369_1997",
    "YEXP-300_1997",
    "YEXP-1500_1997",
    "YEXP-1600_1997",
    "YEXP-1800_1997",
    "YEXP-2000_1997",
    "sex",
    "KEY_BDATE_M_1997",
    "KEY_BDATE_Y_1997",
    "PC8-090_1997",
    "PC8-092_1997",
    "PC9-002_1997",
    "PC12-024_1997",
    "PC12-028_1997",
    "CV_AGE_12/31/96_1997",
    "CV_BIO_MOM_AGE_CHILD1_1997",
    "CV_BIO_MOM_AGE_YOUTH_1997",
    "CV_CITIZENSHIP_1997",
    "CV_ENROLLSTAT_1997",
    "CV_HH_NET_WORTH_P_1997",
    "CV_YTH_REL_HH_CURRENT_1997",
    "CV_MSA_AGE_12_1997",
    "CV_URBAN-RURAL_AGE_12_1997",
    "CV_SAMPLE_TYPE_1997",
    "CV_HGC_BIO_DAD_1997",
    "CV_HGC_BIO_MOM_1997",
    "CV_HGC_RES_DAD_1997",
    "CV_HGC_RES_MOM_1997",
    "race",
    "YSCH-6800_1998",
    "YSCH-7300_1998",
    "YSAQ-372B_1998",
    "YSAQ-371_2000",
    "YSAQ-282J_2002",
    "YSAQ-282Q_2002",
    "CV_HH_NET_WORTH_Y_2003",
    "CV_BA_CREDITS.01_2004",
    "YSAQ-000B_2004",
    "YSAQ-373_2004",
    "YSAQ-369_2005",
    "CV_BIO_CHILD_HH_2007",
    "YTEL-52~000001_2007",
    "YTEL-52~000002_2007",
    "YTEL-52~000003_2007",
    "YTEL-52~000004_2007",
    "CV_BIO_CHILD_HH_2009",
    "CV_COLLEGE_TYPE.01_2011",
    "CV_INCOME_FAMILY_2011",
    "CV_HH_SIZE_2011",
    "CV_HH_UNDER_18_2011",
    "CV_HH_UNDER_6_2011",
    "CV_HIGHEST_DEGREE_1112_2011",
    "CV_BIO_CHILD_HH_2011",
    "YSCH-3112_2011",
    "YSAQ-000A000001_2011",
    "YSAQ-000A000002_2011",
    "YSAQ-000B_2011",
    "YSAQ-360C_2011",
    "YSAQ-364D_2011",
    "YSAQ-371_2011",
    "YSAQ-372CC_2011",
    "YSAQ-373_2011",
    "YSAQ-374_2011",
    "YEMP_INDCODE-2002.01_2011",
    "CV_BIO_CHILD_HH_2015",
    "YEMP_INDCODE-2002.01_2017",
    "YEMP_OCCODE-2002.01_2017",
    "CV_MARSTAT_COLLAPSED_2017",
    "YINC-1400_2017",
    "income",
    "YINC-1800_2017",
    "YINC-2400_2017",
    "YINC-2600_2017",
    "YINC-2700_2017",
    "CVC_YTH_REL_HH_AGE6_YCHR_XRND",
    "CVC_SAT_MATH_SCORE_2007_XRND",
    "CVC_SAT_VERBAL_SCORE_2007_XRND",
    "CVC_ACT_SCORE_2007_XRND",
    "CVC_HH_NET_WORTH_20_XRND",
    "CVC_HH_NET_WORTH_25_XRND",
    "CVC_ASSETS_FINANCIAL_25_XRND",
    "CVC_ASSETS_DEBTS_20_XRND",
    "CVC_HH_NET_WORTH_30_XRND",
    "CVC_HOUSE_VALUE_30_XRND",
    "CVC_HOUSE_TYPE_30_XRND",
    "CVC_ASSETS_FINANCIAL_30_XRND",
    "CVC_ASSETS_DEBTS_30_XRND")

```

## 1a. Data Processing

Create new variable columns for: <br />
    - Employment Industry (emp.industry) <br />
    - Income Levels for 1997 Based on Income Data (income.level_1997) <br />
    - Encarceration History (ever.incarcerated) <br />
    - Biological Relationship of Home Parental Figure (bio.status) <br />
    - Number of Parents within Home (num.parent) <br />

Recode a number of preexisting columns to better support analysis

```{r, echo=FALSE}
#Rename and Recode Select Covariates
nlsy <- as_tibble(nlsy) %>%
  rename(cocaine_use = 'YSAQ-372CC_2011',
         smokes = 'YSAQ-360C_2011',
         drinks = 'YSAQ-364D_2011',
         HHIC_1997 = 'CV_HH_NET_WORTH_P_1997',
         first.incarceration = 'INCARC_AGE_FIRST_XRND',
         residence = 'CV_URBAN-RURAL_AGE_12_1997') %>%
  mutate(marital_status = recode_factor(CV_MARSTAT_COLLAPSED_2017, 
                                        `0`="NEVER-MARRIED", 
                                        `1` = "MARRIED", 
                                        `2` = "SEPARATED", 
                                        `3` = "DIVORCED", 
                                        `4` = "WIDOWED", 
                                        `-5` = "UNKNOWN", 
                                        `-3` = "UNKNOWN")) %>%
  mutate(sex = recode_factor(sex, `1`="MALE", #set Male as Baseline
                             `2`="FEMALE")) %>%
  mutate(race = recode_factor(race, 
                              `4` = "NON-BLACK/NON-HISPANIC",#set non-Black/non-Hispanic as Baseline
                              `1` = "BLACK", 
                              `2` = "HISPANIC", 
                              `3` = "MIXED-RACE (NON-HISP.)"
                              )) %>%
  mutate(HHIC_1997 = case_when(HHIC_1997<0 ~ NA_real_, HHIC_1997>0 ~ HHIC_1997)) %>%
  
  #Create industry column based upon existing YEMP_INDCODE-2002.01_2017 variable
  mutate(emp.industry = cut(`YEMP_INDCODE-2002.01_2017`, breaks = c(-5,0,169,290,369,490,569,690,769,770,1069,3990,4069,4590,4669,5790,5889,5890,6069,
                                                                    6390,6469,6780,6869,7190,7269,7790,7859,8470,8559,8690,8769,9290,9369,9590,9669,9890,9949,9990),
                            labels = c("UNKNOWN","NA","AGRICULTURE, FORESTRY, FISHING","NA","MINING","NA","UTILITIES","NA","CONSTRUCTION","NA","MANUFACTURING","NA",
                                       "WHOLESALE TRADE","NA","RETAIL TRADE","NA","ENTERTAINMENT","NA","TRANSPORTATION AND wHAREHOUSING","NA",
                                       "INFORMATION AND COMMUNICATION", "NA", "FINANCE, INSURANCE, AND REAL ESTATE","NA","PROFESSIONAL AND RELATED SERVICES", "NA",
                                       "EDUCATIONAL, HEALTH, AND SOCIAL SERVICES","NA","ENTERTAINMENT, ACCOMODATIONS, AND FOOD SERVICES", "NA","OTHER
                                       SERVICES","NA","PUBLIC ADMINISTRATION","NA","ACTIVE DUTY MILITARY","NA","ACS SPECIAL CODES"),include.lowest = TRUE)) %>% 
  
  mutate(birth.mom.age = cut(CV_BIO_MOM_AGE_YOUTH_1997, breaks = c(-Inf,0,19,29,39,49,59), 
                             labels = c("Unknown","Child/Teen","20s","30s","40s","50s+"),ordered_result = TRUE)) %>%
  
  mutate(income.level_1997 = cut(HHIC_1997, breaks = c(-Inf, 0, 18502, 37005, 55500, 74010,92512,111015,Inf), 
                                 labels = c("NA", "Under Federal Poverty Level", "Low Income", "Above Median Income", 
                                            "Comfortable", "Upper Middle Class","Wealthy", "Very Wealthy"), ordered_result = TRUE)) %>%
  
  mutate(first.incarceration = case_when(first.incarceration<0 ~ NA_real_, first.incarceration>0 ~ first.incarceration)) %>%
   
  mutate(residence = recode_factor(residence,`2` = "UNKNOWN",
                                   `-3` = "UNKNOWN",
                                   `-4` = "UNKNOWN",
                                   `0` = "RURAL",
                                   `1` = "URBAN")) %>%
  mutate_at(c("cocaine_use", "smokes", "drinks"), ~ recode_factor(.x, `0` = "NO", `1` = "YES", .default = "UNKNOWN")) %>%
  #creates additional column that indicates whether or not an individual has ever been incarcerated
  mutate(ever.incarcerated = case_when(INCARC_TOTNUM_XRND ==0 ~ "NO", INCARC_TOTNUM_XRND >0 ~ "YES")) %>%
  #creates additional column that indicates whether or not child lives with at least one biological parent
  mutate(bio.status = case_when(CVC_YTH_REL_HH_AGE6_YCHR_XRND == 1 ~ "BIO", 
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 2 ~ "BIO",
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 3 ~ "BIO",
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 4 ~ "BIO",
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 5 ~ "BIO",
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 6 ~ "NON.BIO",
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 7 ~ "NON.BIO",
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 8 ~ "NON.BIO",
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 9 ~ "NON.BIO",
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 10 ~ "NON.BIO",
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == -3 ~ "UNKNOWN",
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == -4 ~ "UNKNOWN")) %>%
  #creates additional column that indicates Number of parents in the household (0,1,2)
  #Child assigned to at least one parent unless explicitly stated living in group quarters
  mutate(num.parent = case_when(CVC_YTH_REL_HH_AGE6_YCHR_XRND == 1 ~ 2, 
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 2 ~ 2,
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 3 ~ 2,
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 4 ~ 1,
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 5 ~ 1,
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 6 ~ 2,
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 7 ~ 2,
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 8 ~ 1,
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 9 ~ 0,
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == 10 ~ 0,
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == -3 ~ NA_real_,
                                CVC_YTH_REL_HH_AGE6_YCHR_XRND == -4 ~ NA_real_))

```

## 1b. Data Summarization: Basic Data Summary of Key Variables

### THE MOTIVATION: Gender Wage Gap
```{r, echo=FALSE}
t.test.wage.gap <- t.test(income ~ sex, data = nlsy) #statistically significant difference between avg incomes for men and women

income.data.gender <- nlsy %>%
  group_by(sex) %>%
 summarize(avg.income = mean(income),
           upper = t.test(income)$conf.int[1],
           lower = t.test(income)$conf.int[2]
           )

ggplot(data = income.data.gender, aes(x=sex, y=avg.income, fill=I("steelblue"))) + geom_bar(stat = "identity") +
  xlab("Gender") +
  ylab("Income") +
  ggtitle("AVG Income of Men and Women") +
  guides(fill = FALSE) +
  geom_errorbar(aes(ymax = upper, ymin = lower), width = 0.1, size = 1)
  #obvious gap between average income between men and women. Appears to be statistically significant based off of error bars
```

<font color="#157515"><b>


> Here we can see both in the T-Test output as well as the bar chart that there is a statistically significant wage gap between genders. On average men make  $`r round ((t.test.wage.gap$estimate[1]-t.test.wage.gap$estimate[2]), 2) ` more than women.

</font></b>

### THE BASICS: Data Summary By Category
We identified at least two key variables for each major category, which we
felt to have a high chance of influence income outcomes between gender

### Demographic Information:
```{r, echo=FALSE}
kable(table(nlsy$sex), col.names = c("Gender Designation", "Count"))
```
<font color="#157515"><b>

>The data is nearly evently split between genders.

NOTE: we removed active duty military because there were no women in the survey and could not calculate a wage gap

</font></b>
```{r, echo=FALSE}
kable(table(nlsy$race), col.names = c("Race Designation", "Count"))
```

<font color="#157515"><b>

>The data is based towards  non-black/non-hispanic individuals, but the quantities for black and hispanic participants
are well over 1000, which indicates that for most analyses we should have sufficient data to make statistical inferences.
Mixed-raced (non-hispanic) individuals are poorly represented in the data.  All considerations regarding this demographic should be closely considered, especially
for subsets of data where N < 32.

NOTE: we removed active duty military because there were no women in the survey and could not calculate a wage gap

</font></b>

### Geographical Information:
```{r, echo=FALSE}
kable(table(nlsy$emp.industry), col.names = c("Industry", "Count"))
```

<font color="#157515"><b>

>A significant portion of individuals are in "UNKNOWN" industries, reducing the overall quantity of discernable industries accross which do perform
analyses.  A number of industries have < 100 participants, which may bode poorly for future analysis on this variable.

NOTE: we removed active duty military because there were no women in the survey and could not calculate a wage gap

</font></b>

### Key Question: Does the gender wage gap persist over industries? Does it increase or decrease according to industry?
```{r, echo = FALSE}
nlsy.no.military <- subset(nlsy, emp.industry!="ACTIVE DUTY MILITARY")

gap.data.gender.industry <- nlsy.no.military %>% 
  group_by(emp.industry) %>%
  summarize(income.gap = mean(income[sex == "MALE"], na.rm = TRUE) -
              mean(income[sex == "FEMALE"], na.rm = TRUE),
           upper = t.test(income ~ sex)$conf.int[1],
           lower = t.test(income ~ sex)$conf.int[2],
           is.significant = as.numeric(t.test(income ~ sex)$p.value < 0.05))

gap.data.gender.industry$is.significant <- recode(gap.data.gender.industry$is.significant, `0` = FALSE, `1` = TRUE)

ggplot(data = gap.data.gender.industry,aes(x=emp.industry, y=income.gap, fill=is.significant)) + 
  geom_bar(stat = "identity") +
  xlab("Industry") +
  ylab("Income Gap ($)") +
  ggtitle("Income Gap between Men and Women, by Industry") +
  theme(axis.text.x=element_text(angle = -90, vjust = 0, size=5))
```
<font color="#157515"><b>

>Persistently statistically significant wage gap between men and women even across most industries, with Agriculture/Forestry/Fishing and Finance/Insurance/Real Estate having the largest wage gap. Because there are large differences in the wage gaps betindustries we will include it in our model. 

NOTE: we removed active duty military because there were no women in the survey and could not calculate a wage gap

</font></b>

```{r, echo=FALSE}
kable(table(nlsy$residence), col.names = c("Location of Home", "Count"))
```

### Key Question: Does the gender wage gap persist accross location of residence?
```{r, echo = FALSE}

##Residence: URban and Rural
gap.data.residence <- nlsy %>%
  group_by(residence) %>%
  summarize(income.gap = (mean(income[sex == "MALE"], na.rm = TRUE) - mean(income[sex == "FEMALE"], na.rm = TRUE)),
              'lower' = t.test(income ~ sex)$conf.int[1],
              'upper' = t.test(income ~ sex)$conf.int[2])
#No statistical difference between income gap between men and women in urban and rural environments
#There is a larger spread in income difference for rural communities, which may be a function
#of a generally smaller population of people who identify as being in a rural residence
ggplot(subset(gap.data.residence, residence %in% c("RURAL", "URBAN")), 
       aes(x=residence, y=income.gap, fill=I("steelblue"))) + geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, position = position_dodge(0.9)) +
  xlab("Location of Residence") +
  ylab("Income Gap($)") +
  ggtitle("Income Gap between Men and Women, by Residence Location") +
  guides(fill = FALSE)


```
<font color="#157515"><b>

>There is not a statistically significant difference in income gap accross urban and rural regions, but it does appear that
amongst people who live in rural areas the spread of the income gap is larger.  A potential influencing factor could be a larger
disparity amongst types of jobs (and subsequently their respective salaries) in rural regions as compared to urban regions.

</font></b>

### Home Demographics Information:
```{r, echo=FALSE}
kable(table(nlsy$marital_status), col.names = c("Marital  Status", "Count"))
kable(table(nlsy$birth.mom.age), col.names = c("Age of Mother", "Count"))
```
```{r, echo=FALSE}
kable(table(nlsy$income.level_1997), col.names = c("Income Bracket", "Count"))
```

<font color="#157515"><b>

>For both Marital Status and Income Bracket variables we see a healthy spread of participants, except  widowed and above 40 year old age brackets.

</font></b>

```{r, echo=FALSE}
#Average net worth of household in 1997 by gender as reported by parent

#Filter top-coded data
nlsy.filtered <- filter(nlsy,HHIC_1997 < 600000 & income < 200000 & !is.na(nlsy$sex))

income.data.HHIC1997 <- nlsy.filtered %>%
  group_by(sex) %>%
  summarize("avg.HHIC_1997" = mean(HHIC_1997, na.rm = TRUE),
            'lower' = t.test(HHIC_1997, mu = mean(HHIC_1997, na.rm = TRUE))$conf.int[1],
            'upper' = t.test(HHIC_1997, mu = mean(HHIC_1997,na.rm = TRUE))$conf.int[2])

ggplot(data = income.data.HHIC1997, aes(x=sex, y=avg.HHIC_1997, fill=I("steelblue"))) + 
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, position = position_dodge(0.9)) +
  xlab("Gender Designation") +
  ylab("AVG Household Net Worth 1997") +
  ggtitle("AVG Household Net Worth 1997 of Men and Women (According to Parent)") +
  guides(fill = FALSE) 

```

<font color="#157515"><b>

>There is not a statistically significant difference in average Household net worth between male and female children in 1997, which was an important check, given
the known impact of living conditions on outomes for child success.

NOTE: Top-coded data (net worth above $599,99.99) were eliminated in order to remove biased created by ungranular top-coding function .
</font></b>



```{r, echo=FALSE}

#No significant difference in HHIC_1997 between men and women
#amongst the same demographic, but you do observie a significant right tale:
#the data is not normally distributed, which shows that white individuals
#have HHIC_1997 > than other groups
ggplot(data = nlsy.filtered, aes(x=HHIC_1997)) +
  geom_density(alpha = 0.5, aes(fill = sex)) +
  xlab("Household Net Worth 1997 ($)") +
  ylab("Count") +
  ggtitle("Density Plots of Household Net Worth per Gender") +
  guides(fill = FALSE) 

```

<font color="#157515"><b>

>Further plotting of the household net worth data confirmed similar trends between male and female populations.  Plotting futher confirmed that net worth distributions amongst male and female children in 1997 are not normal.  Any further analysis utilizing this variable that assumes normalty will require close consideration and/or some form of transformation.

</font></b>


### Drug Use & Incarceration Information:
```{r, echo=FALSE}
kable(table(nlsy$cocaine_use), col.names = c("Cocaine Use Status", "Count"))
kable(table(nlsy$smokes), col.names = c("Smoking Use Status", "Count"))
kable(table(nlsy$drinks), col.names = c("Drinking Use Status", "Count"))
```

```{r, echo = FALSE}
#Investigation on Incarceration
nlsy.filtered <- filter(nlsy, income < 200000 & !is.na(nlsy$sex))
incarceration.age.data <- nlsy %>%
  group_by(sex) %>%
  summarize("avg.firstincarcerate" = mean(first.incarceration, na.rm = TRUE),
            'lower' = t.test(first.incarceration, mu = mean(first.incarceration, na.rm = TRUE))$conf.int[1],
            'upper' = t.test(first.incarceration, mu = mean(first.incarceration,na.rm = TRUE))$conf.int[2])

ggplot(data = incarceration.age.data, aes(x=sex, y=avg.firstincarcerate, fill=I("steelblue"))) + 
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, position = position_dodge(0.9)) +
  xlab("Gender") +
  ylab("Average Age of Incarceration") +
  ggtitle("Average Age of Incarceration (According to Gender)") +
  guides(fill = FALSE)

#Men on average have a slightly higher average incarceration age than women.
#Data potentially look statistically significant.
```

```{r, echo = FALSE}

#Data seems conglomerated around specific age ranges
#Not enough women incarcerated
#Does look like income is not as high for those incarcerated.
#A lot of zero incomes for incarceration at any age
ggplot(data = nlsy.filtered, aes(x=first.incarceration, y=income, colour = sex, shape = sex))+
  geom_point(alpha = 0.3) +
  xlab("Age at First Incarceration (yrs)") +
  ylab("Income ($)") +
  ggtitle("Income vs. Age of First Incarceration") +
  guides(fill = FALSE)

```

```{r, echo = FALSE}

#Output table to test whether odds of having above zero income for previously incarcerated women \
# is lower than previously incarcerated men 
incarceration.income.table <- filter(nlsy, nlsy$ever.incarcerated == 'YES') %>%
  mutate(non.zincome = case_when(income <=0 ~ "NO", income >0 ~ "YES"))
incarceration.income.table <- with(incarceration.income.table, table(non.zincome, sex))

#The odds of making non-zero income for men who have been incarcerated
# is 1.43266 higher than for incarerated women.
incarceration.fisher.test <- fisher.test(incarceration.income.table)

```


## 2. Methodology
### What factors in particular play a role the gender wage gap?
We chose to explore several subsets of data before running linear models.  
Our intent was to identify any interesting patterns prior to running our analyses.

We started with race and marital status and then furthered the analysis with a reassessment of 
household net worth, and we concluded with incarceration data and income

All of these variables seemed the most impactful from an intuitive perspective.

### Race and Marital Status

```{r}
gap.data.gender.race <- nlsy %>% 
  group_by(race) %>%
  summarize(income.gap = mean(income[sex == "MALE"], na.rm = TRUE) 
            - mean(income[sex == "FEMALE"], na.rm = TRUE),
           upper = t.test(income ~ sex)$conf.int[1],
           lower = t.test(income ~ sex)$conf.int[2],
           is.significant = as.numeric(t.test(income ~ sex)$p.value < 0.05))
  
gap.data.gender.race$is.significant <- recode(gap.data.gender.race$is.significant, `0` = FALSE, `1` = TRUE)


ggplot(data = gap.data.gender.race, aes(x=race, y=income.gap, fill=is.significant)) + geom_bar(stat = "identity") +
  xlab("Race") +
  ylab("Income Gap($)") +
  ggtitle("Income Gap between Men and Women, by Race") +
    theme(axis.text.x=element_text(angle = -15, vjust = 0)) +
    geom_errorbar(aes(ymax = upper, ymin = lower), width = 0.1, size = 1)
```
<font color="#157515"><b>

>Income gap does persist over race but there is noticeable difference in the gaps.
The wage gap is quite large and statistically significant for hispanic men and women and for non-black non-hispanic men and women. However, I would pause before concluding this test shows that wage gap between black men and women or mixed race men and women is insignificant. There might be some counfounding variables in the background. Black people do generally have lower wages than white people.

</font></b>

```{r, echo = FALSE}
nlsy.valid.mar_stat <- subset(nlsy, marital_status != "UNKNOWN")
gap.data.marital_stat <- nlsy.valid.mar_stat %>%
  group_by(marital_status) %>%
 summarize(income.gap = mean(income[sex == "MALE"], na.rm = TRUE) 
           - mean(income[sex == "FEMALE"], na.rm = TRUE),
           upper = t.test(income ~ sex)$conf.int[1],
           lower = t.test(income ~ sex)$conf.int[2],
           is.significant = as.numeric(t.test(income ~ sex)$p.value < 0.05))

gap.data.marital_stat$is.significant <- recode(gap.data.marital_stat$is.significant, `0` = FALSE, `1` = TRUE)

ggplot(data = gap.data.marital_stat, aes(x=marital_status, y=income.gap, fill=is.significant)) + geom_bar(stat = "identity") +
  xlab("Marital Status") +
  ylab("Income Gap($)") +
  ggtitle("Income Gap between Men and Women, by Marital Status") +
    theme(axis.text.x=element_text(angle = -90, vjust = 0)) +
  geom_errorbar(aes(ymax = upper, ymin = lower), width = 0.1, size = 1)

```
<font color="#157515"><b>

>The wage gap does persist over marital status. In particular, married men and women seem to have the largest disparity in incomes. This is not entirely surprising given that traditional gender roles are still adhered to any many parts of the country. If the man is going to be the only one earning an income then it would follow that there would be a large disparity in wages. It's worth noting we chose to subset the nlsy data to only include those whose marital status was known. We removed from the original data set.

>These finding agree with our original intuition that race and marital status do have a significant effect on the gender wage gap. 

</font></b>

### Substance Use: Alcohol, Tobacco, Hard Drugs(Cocaine)
```{r, echo = FALSE}
### ALCOHOL, TOBACCO, HARD DRUGS(Cocaine)
nlsy.valid_smokes <- subset(nlsy, smokes!= "UNKNOWN", select = c(income,sex, smokes))

gap.data.gender.smokes <- nlsy.valid_smokes %>% 
  group_by(smokes) %>%
  summarize(income.gap = mean(income[sex == "MALE"], na.rm = TRUE) - mean(income[sex == "FEMALE"], na.rm = TRUE),
           upper = t.test(income ~ sex)$conf.int[1],
           lower = t.test(income ~ sex)$conf.int[2],
           is.significant = as.numeric(t.test(income ~ sex)$p.value < 0.05))

gap.data.gender.smokes$is.significant <- recode(gap.data.gender.smokes$is.significant, `0` = FALSE, `1` = TRUE)

ggplot(data = gap.data.gender.smokes, aes(x=smokes, y=income.gap, fill=is.significant)) + geom_bar(stat = "identity") +
  xlab("Smokes") +
  ylab("Income Gap($)") +
  ggtitle("Income Gap between Men and Women, by Smokes") +
  geom_errorbar(aes(ymax = upper, ymin = lower), width = 0.1, size = 1)


nlsy.valid_cocaine <- subset(nlsy, cocaine_use != "UNKNOWN", select = c(income,sex, cocaine_use))

gap.data.gender.cocaine_use <- nlsy.valid_cocaine %>% 
  group_by(cocaine_use) %>%
  summarize(income.gap = mean(income[sex == "MALE"], na.rm = TRUE) - mean(income[sex == "FEMALE"], na.rm = TRUE),
           upper = t.test(income ~ sex)$conf.int[1],
           lower = t.test(income ~ sex)$conf.int[2],
           is.significant = as.numeric(t.test(income ~ sex)$p.value < 0.05))

gap.data.gender.cocaine_use$is.significant <- recode(gap.data.gender.cocaine_use$is.significant, `0` = FALSE, `1` = TRUE)

ggplot(data = gap.data.gender.cocaine_use, aes(x=cocaine_use, y=income.gap, fill=is.significant)) + geom_bar(stat = "identity") +
  xlab("Cocaine Use") +
  ylab("Income Gap($)") +
  ggtitle("Income Gap between Men and Women, by Cocaine Use") +
    geom_errorbar(aes(ymax = upper, ymin = lower), width = 0.1, size = 1)


nlsy.valid_drinks<- subset(nlsy, drinks != "UNKNOWN", select = c(income,sex,drinks))

gap.data.gender.drinks <- nlsy.valid_drinks %>% 
  group_by(drinks) %>%
  summarize(income.gap = mean(income[sex == "MALE"], na.rm = TRUE) -
              mean(income[sex == "FEMALE"], na.rm = TRUE),
           upper = t.test(income ~ sex)$conf.int[1],
           lower = t.test(income ~ sex)$conf.int[2],
           is.significant = as.numeric(t.test(income ~ sex)$p.value < 0.05))

gap.data.gender.drinks$is.significant <- recode(gap.data.gender.drinks$is.significant, `0` = FALSE, `1` = TRUE)


ggplot(data = gap.data.gender.drinks, aes(x=drinks, y=income.gap, fill=is.significant)) + geom_bar(stat = "identity") +
  xlab("Drinks") +
  ylab("Income Gap($)") +
  ggtitle("Income Gap between Men and Women, by Drinks") +
    geom_errorbar(aes(ymax = upper, ymin = lower), width = 0.1, size = 1)

```
<font color="#157515"><b>
>The statistically significant wage gap persists over those who drink, smoke, and do hard drugs. This suggests that substance use has no affect on the wage gap.
There is a noticeable difference in the wage gap between men and women who drink and don't drink is considerable but still not statistically significant given the error bars overlap. We will not include these variables in our linear model.

</font></b>

### Revisiting Houshold Income 
```{r,  echo = FALSE}
#No sttistically significant difference in average houshold net worth between men and women

#you see a less of a range between 150K and mx income which is interesting, there is not gradient like with 
#Percentage of high earning men is clearly greater
ggplot(data = nlsy.filtered, aes(x=HHIC_1997, y=income, colour = sex, shape = sex))+
  geom_point(alpha = 0.3)
```

<font color="#157515"><b>

>Plotting household networth against income shows a heavy clustering of data in the sub-$75000 income and sub- $200000 net worth region of the graph.
This could appears to be more a function of the demographics of the population surveyed, given prior summary and the fact that a large range of incomes
is can be associated with any an equally large range of household net worths.

</font></b>

### Now we'll explore the relationship between mother's age when child is born and income
    - Is there a relationship between age of mother and the wage gap?
```{r}
nlsy.valid_mom_age <- subset(nlsy, CV_BIO_MOM_AGE_YOUTH_1997 < 55 & CV_BIO_MOM_AGE_YOUTH_1997 > 0)
nlsy.valid_mom_age <- subset(nlsy.valid_mom_age, income < 200000)

gap.data.AGE_MOM_YOUTH <- nlsy.valid_mom_age %>%
  group_by(birth.mom.age) %>%
 summarize(
   income.gap = mean(income[sex == "MALE"], na.rm = TRUE) -
              mean(income[sex == "FEMALE"], na.rm = TRUE)
   )

ggplot(data = gap.data.AGE_MOM_YOUTH, aes(x=birth.mom.age, y=income.gap, fill=I("steelblue"))) + geom_bar(stat = "identity") +
  xlab("Age of Mother at Birth of Child") +
  ylab("Income Gap($)") +
  ggtitle("Income Gap between Men and Women, by Age of Mother at Birth") +
  guides(fill = FALSE)
```

<font color="#157515"><b>

>Here we see that the income gap persists over different age at birth categories of mother. Suggesting that age of birth mother does not have an effect on the wage gap.

</font></b>

### Is mother's age at all associated with income?
Intuitively, we would expect mother's age at birth of child to be positively correlated with income. Generally younger mother's have lower incomes.
```{r, echo = FALSE}
plot.interact <- qplot(x = CV_BIO_MOM_AGE_YOUTH_1997, y = income, color = sex, data = nlsy.valid_mom_age) + stat_smooth(method = "lm", se = FALSE, fullrange = TRUE)

plot.interact

```
<font color="#157515"><b>

>Here we can see that mother's age when child was born is modestly positively correlated with income for both men and women. Because this positive correlation exists equally for men and women, it likely does not affect the income gap and we will not include it in our model.

Note: we did subset the data in this case to remove rows if they had invalid or unknown values for age of mother when she gave birth. We also removed the top coded values of income to increase the accuracy of our model.

</font></b>

#### Is mother's age at all associated with race?
```{r, echo = FALSE}
ggplot(nlsy.valid_mom_age, aes(x = CV_BIO_MOM_AGE_YOUTH_1997,)) + xlab("Mother's Age at Birth of Child") + geom_histogram(aes(fill = race))

```

<font color="#157515"><b>

>Seemingly no from this data set. Birth mother's age is distributed similarly for every race.

</font></b>

### Collinearity Testing -12/12/20

```{r, echo = FALSE}
### COLINEARITY TESTS

batch.one.var.names <-c("residence","emp.industry")

batch.two.var.names <-c("cocaine_use", "smokes", "drinks", "first.incarceration")

batch.three.var.names <-c("first.incarceration", "CV_BIO_MOM_AGE_YOUTH_1997")

batch.four.var.names <- c("HHIC_1997","CV_BIO_MOM_AGE_YOUTH_1997","num.parent", "CV_BIO_CHILD_HH_2007" )

batch.five.var.names <- c("HHIC_1997","CV_BIO_MOM_AGE_YOUTH_1997", "residence","bio.status", "CV_BIO_CHILD_HH_2007" )

ggpairs(nlsy[,c(batch.three.var.names)], axisLabels = "internal", cardinality_threshold = 20)

ggpairs(nlsy[,c(batch.four.var.names)], axisLabels = "internal", cardinality_threshold = 20)

ggpairs(nlsy[,c(batch.five.var.names)], axisLabels = "internal", cardinality_threshold = 20)

```


## 3. Findings
```{r, echo = FALSE}
###LET'S RUN SOME REGRESSIONS

lm.sex <- lm(income ~ sex, data = nlsy)
kable(coef(summary(lm.sex)), digits = c(0, 0, 2, 4))

lm.race.sex <- lm(income ~ sex*race, data = nlsy)
kable(coef(summary(lm.race.sex)), digits = c(0, 0, 2, 4))

lm.sex.mar_stat <- lm(income ~ sex*marital_status, data = nlsy)
kable(coef(summary(lm.sex.mar_stat)), digits = c(0, 0, 2, 4))

lm.race.sex.residence <-lm(income ~ sex*race*residence, data = nlsy)
#kable(coef(summary(lm.race.sex.residence)), digits = c(0, 0, 2, 4))

lm.race.sex.mar_stat <- lm(income ~ sex*race*marital_status, data = nlsy)
#kable(coef(summary(lm.race.sex.mar_stat)), digits = c(0, 0, 2, 4))

lm.race.sex.industry <- lm(income ~ sex*race*emp.industry, data = nlsy)
#kable(coef(summary(lm.race.sex.industry)), digits = c(0, 0, 2, 4))

lm.race.sex.mar_stat.industry <- lm(income ~ sex*race*marital_status*emp.industry, data = nlsy)
#kable(coef(summary(lm.race.sex.mar_stat.industry)), digits = c(0, 0, 2, 4))

anova(lm.sex, lm.race.sex, lm.sex.mar_stat, lm.race.sex.mar_stat, lm.race.sex.residence, lm.race.sex.industry, lm.race.sex.mar_stat.industry)

```
